A High Efficiency MAC Protocol for WLANs: Providing Fairness in Dense Scenarios

### 2015-10-29 journal.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

\boldmath Collisions are a main cause of throughput degradation in WLANs. The current contention mechanism used in IEEE 802.11 networks is called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA). It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability. Nevertheless, CSMA/CA relies on a random backoff that while effective and fully distributed, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel. Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a fully distributed manner using a deterministic backoff after successful transmissions. Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule. CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness.

\textcolor{red}{This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes. Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors over collision-free schedules. Finally, experimental results on throughput and lost frames from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
\end{abstract}

\begin{IEEEkeywords}
CSMA/ECA, WLAN, MAC, Collision-free, \textcolor{red}{Implementation}.
\end{IEEEkeywords}

@@

A CSMA/CA replacement should be able to provide advantages in terms of throughput, spectrum efficiency and number of supported contenders. All of the aforementioned without sacrificing short-term throughput fairness. Furthermore, it must also serve existing users, which means it has to be backwards compatible.

A suitable candidate, and the one to be evaluated in this work, is called Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)~\cite{barcelo2008lba}. It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism. In CSMA/ECA, nodes use a deterministic backoff after successful transmissions, constructing a collision-free schedule among successful contenders in a fully decentralised way. This backoff mechanism ensures that more channel time is spent on successful transmissions rather than recovering from collisions, thus increasing the throughput of the network. Further enhancements (or extensions), like \emph{Hysteresis} and \emph{Fair Share}~\cite{research2standards} allow CSMA/ECA to support many more contenders in a collision-free schedule. Moreover, CSMA/ECA and its extensions are designed for allowing nodes to transmit as frequently as possible while keeping an even distribution of the available bandwidth among users.

The 802.11 High Efficiency WLAN (HEW) Study Group (SG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios}, specifically those usually encountered at stadiums or conference rooms. Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount. Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement. Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users. Furthermore, the impact of channel errors and imperfect clocks over the deterministic backoff mechanism is also lacking.

@@

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item \textcolor{red}{The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.}
	\textcolor{red}{\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.}
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
\end{itemize}

@@

An overview of similar distributed and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}. CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}. Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results. An overview of CSMA/ECA real-hardware implementation and testbed description is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}. Conclusions are drawn in Section~\ref{conclusions}.

@@

\section{Conclusions}\label{conclusions}
CSMA/ECA$_{\text{Hys+FS}}$ is able to construct a collision-free schedule with many contenders. Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes. All of these while preserving throughput f	airness. Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.

@@

To top it all, the real hardware implementation of CSMA/ECA$_{\text{Hys}}$ with Schedule Reset clearly outperforms CSMA/CA. In itself, the prototype represents a strong indicator that the proposed enhanced collision avoidance could be considered as a viable replacement or evolution of the current MAC for WiFi.

>>>>

\boldmath Collisions are a main cause of throughput degradation in WLANs. The current contention mechanism used in IEEE 802.11 networks is called Carrier Sense Multiple Access with Collision Avoidance (CSMA/CA). It uses a Binary Exponential Backoff (BEB) technique to randomise each contender attempt of transmitting, effectively reducing the collision probability. Nevertheless, CSMA/CA relies on a random backoff that while effective and fully decentralised, in principle is unable to completely eliminate collisions, therefore degrading the network throughput as more contenders attempt to share the channel.

To overcome these situations, Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA) is able to create a collision-free schedule in a fully decentralised manner using a deterministic backoff after successful transmissions. Hysteresis and Fair Share are two extensions of CSMA/ECA to support a large number of contenders in a collision-free schedule. CSMA/ECA offers better throughput than CSMA/CA and short-term throughput fairness. \textcolor{red}{This work describes CSMA/ECA and its extensions. Additionally, it provides the first evaluation results of CSMA/ECA with non-saturated traffic, channel errors, and its performance when coexisting with CSMA/CA nodes. Furthermore, it describes the effects of imperfect clocks over CSMA/ECA and present a mechanism to leverage the impact of channel errors and the addition/withdrawal of nodes over collision-free schedules. Finally, experimental results on throughput and lost frames from a CSMA/ECA implementation using commercial hardware and open-source firmware are presented.}
\end{abstract}

\begin{IEEEkeywords}
CSMA/ECA, WLAN, MAC, Collision-free, \textcolor{red}{Testbed}.
\end{IEEEkeywords}

@@

A CSMA/CA replacement should be able to provide advantages in terms of throughput, spectrum efficiency and number of supported contenders. All of the aforementioned without sacrificing short-term throughput fairness. Furthermore, WLANs implementing such a replacement must also serve existing users, which means they have to be backwards compatible.

A suitable candidate, and the one to be evaluated in this work, is called Carrier Sense Multiple Access with Enhanced Collision Avoidance (CSMA/ECA)~\cite{barcelo2008lba}. It is capable of attaining higher throughput than CSMA/CA by making a simple modification to the contention mechanism, thus keeping maintaining compatibility. In CSMA/ECA, nodes use a deterministic backoff after successful transmissions, constructing a collision-free schedule among successful contenders in a fully decentralised way. This backoff mechanism ensures that more channel time is spent on successful transmissions rather than recovering from collisions, thus increasing the throughput of the network. Further enhancements (or extensions), like \emph{Hysteresis} and \emph{Fair Share}~\cite{research2standards} allow CSMA/ECA to support many more contenders in a collision-free schedule. Moreover, CSMA/ECA and its extensions are designed for allowing nodes to transmit as frequently as possible while keeping an even distribution of the available bandwidth among users.

The 802.11 High Efficiency WLAN (HEW) Task Group (TG) envisions very crowded scenarios as one of the future challenges for WLAN protocols~\cite{HEW-scenarios,bellalta2015WCM}, specifically those usually encountered at stadiums or conference rooms. Further, if the need for serving many users is combined with the increased throughput demands from the upper layers, a performance improvement at the MAC becomes paramount. Although many studies have been made analysing the performance of CSMA/ECA~\cite{barcelo2008lba,research2standards,bellalta2009vtc,E2CA_performance}, there are still several open aspects that require further attention and additional insight to consider CSMA/ECA as a potential CSMA/CA replacement. Namely, neither assesses the protocol's backwards compatibility property or its behavior under non-saturated traffic conditions while serving many users. Furthermore, the impact of channel errors, node addition/withdrawal, and imperfect clocks over the deterministic backoff mechanism is also lacking.

@@

\begin{itemize}
	\item First results on the achievable throughput and delay of CSMA/ECA with Hysteresis and Fair Share under non-saturated traffic conditions for very large number of nodes.
	\item \textcolor{red}{The impact of imperfect clocks and channel errors on CSMA/ECA with Hysteresis and Fair Share's deterministic backoff and its consequences on the achieved performance.}
	\item Formulation of the throughput bounds for CSMA/ECA with Hysteresis and Fair Share.
	\textcolor{red}{\item Introduce the Schedule Reset mechanism for reducing the schedule length in case of users withdrawing from the contention (non-saturated traffic), or when its value is increased due to channel errors.}
	\item Coexistence and backwards compatibility with CSMA/CA nodes under different traffic conditions.
	\item First implementation of CSMA/ECA with Hysteresis in real hardware and the experimental results in a crowded WLAN testbed.
\end{itemize}

@@

An overview of similar decentralised and collision-free MAC protocols for WLANs is provided in Section~\ref{relatedWork}. CSMA/ECA, as well as its extensions for allocating many contenders in a collision-free schedule are explained in Section~\ref{introProtocol}. Section~\ref{simulations} details the simulation environment for testing CSMA/ECA, while Section~\ref{results} explains the results. An overview of CSMA/ECA real-hardware implementation and testbed description is compiled in Section~\ref{EDCA}, followed by a summary of the still missing CSMA/ECA features needed to become the next CSMA/CA replacement in Section~\ref{ECAtoCA}. Conclusions are drawn in Section~\ref{conclusions}.

@@

\section{Conclusions}\label{conclusions}
CSMA/ECA$_{\text{Hys+FS}}$ is able to construct a collision-free schedule with many contenders. Taking advantage of this condition, the cumulative throughput experienced by CSMA/ECA$_{\text{Hys+FS}}$ nodes goes beyond the achievable by CSMA/CA for any number of nodes. All of these while preserving throughput fairness. Further, as non-saturated CSMA/ECA$_{\text{Hys+FS}}$ networks get crowded the cumulative throughput keeps increasing up to the saturation point, contrary to the throughput degradation seen in CSMA/CA networks.

@@

To top it all, the real hardware implementation of CSMA/ECA$_{\text{Hys}}$ clearly outperforms CSMA/CA. In itself, the prototype represents a strong indicator that the proposed enhanced collision avoidance could be considered as a viable replacement or evolution of the current MAC for WiFi.

### 2015-10-29 realImplementation-Gringoli.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

We confirmed the simulations results experimentally with a testbed of real prototypes. This phase was mandatory in order to check whether the behaviour of the system is still correct in the presence of unexpected phenomena, like excessive channel errors, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of the Schedule Reset mechanism with a real time horizon.

@@

For the Schedule Reset mechanism, we added a bitmap for monitoring the state of every single slot after a successful transmission. To index the current slot in the bitmap we used the hardware register {\tt SPR\_IFS\_BKOFFDELAY}, which counts how many slots have still to come before the next transmission. The value of the register is decremented once per idle slot, so the idea is to start with an empty bitmap and mark those slots for which the countdown procedure was interrupted.

@@

We performed four experiments for every number of contenders, starting from 1 up to 25. For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also functions as iPerf server for each flow. We used WiFi channel 14  in order to avoid interference from other networks. The aggregated throughput is derived from the iPerf logs, while the percentage of lost frames is reported by the firmware. This is known to be $(\text{tx}-\text{sx})/\text{tx}$; where tx are the number of transmission attempts, and sx are the number of acknowledged frames.

@@

Nevertheless, the implementation results reveal that there are other underlying factors that disrupt collision-free schedules. This is evidenced by the increasing percentage of lost frames followed by a throughput degradation in CSMA/ECA$_{\text{Hys}}$. The analysis of the factors that may disrupt collision-free schedules on real hardware implementations of CSMA/ECA$_{\text{Hys}}$ is left as a future research topic.

>>>>

We confirmed the simulations results experimentally with a testbed of real prototypes. This phase was mandatory in order to check whether the behaviour of the system is still correct in the presence of real phenomena, like excessive channel errors, or technical limitations of the hardware like imperfect node synchronisation; and to verify the implementation feasibility of CSMA/ECA$_{\text{Hys}}$ and the Schedule Reset mechanism with a real time horizon.

@@

For the Schedule Reset mechanism, we added a bitmap for monitoring the state of every single slot after a successful transmission. To index the current slot in the bitmap we used the hardware register {\tt SPR\_IFS\_BKOFFDELAY}, which counts how many slots have still to come before the next transmission. The value of the register is decremented once per idle slot.

@@

We performed four experiments for each tested number of contenders, starting from 1 and up to 25. For every experiment each station establishes an iPerf~\cite{tirumala2005iperf} session and transmits saturated UDP traffic towards a central AP, which also functions as iPerf server for each flow. We used WiFi channel 14  in order to avoid interference from other networks. The aggregated throughput is derived from the iPerf logs, while the percentage of lost frames is reported by the firmware. This is known to be $(\text{tx}-\text{sx})/\text{tx}$; where tx are the number of transmission attempts, and sx are the number of acknowledged frames.

@@

Nevertheless, the implementation results reveal that there are other underlying factors that disrupt collision-free schedules. This is evidenced by the increasing percentage of lost frames followed by a throughput degradation in CSMA/ECA$_{\text{Hys}}$. The analysis of the factors that may disrupt collision-free schedules on real hardware implementations of CSMA/ECA$_{\text{Hys}}$ is left as future research.

### 2015-10-29 relatedWork.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

Recent advances in the WLANs PHY~\cite{perahia2008ieee,6191306} push the research community towards the development of MAC protocols able to take advantage of a much faster PHY. By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput. Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a fully distributed fashion in order to avoid injecting extra control traffic that may reduce the data throughput.

\textcolor{red}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like ATM~\cite{frigon2001dynamic}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, distributed and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques.}

@@

Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully distributed way. It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $M$-slots in length. Backlogged stations pick a slot in the virtual cycle to attempt transmission. If two or more stations picked the same slot in the cycle, their transmissions will eventually collide. This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided. When all $N$ stations reserve a different slot, a collision-free schedule is achieved.

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $N>M$. Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance\textcolor{red}{\footnote{\textcolor{red}{With a very large $M$ and $N\ll M$, the overall aggregated throughput will start to decrease as $N\rightarrow M$. This is due to the very big period of time between the successful transmissions of each node. For the opposite case, that is $N>M$, the throughput degradation will be accelerated due to collisions.}}}. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

@@

Learning-MAC~\cite{L_MAC} is another MAC protocol able to build a collision-free schedule for many contenders. It does so defining a \emph{learning strength} parameter, $\beta\in(0,1)$. Each contender starts by picking a slot for transmission $s$ of the schedule $n$ of length $C$ at random with uniform probability. After a contender picks slot $s(n)$, its selection in the next schedule, $s(n+1)$, will be conditioned by the result of the current attempt. (\ref{success}) and (\ref{collisions-eq}) extracted from~\cite{L_MAC} show the probability of selecting the same slot $s(n)$ in cycle $n+1$.

@@

L-MAC is able to achieve higher throughput than the current MAC with a very fast convergence speed. Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to the current MAC or may require a centralised entity~\cite{barcelo2011tcf}.

>>>>

Recent advances in the WLANs PHY~\cite{perahia2008ieee,6191306} push the research community towards the development of MAC protocols able to take advantage of a much faster PHY. By reducing the time spent in collisions nodes are able to transmit more often, which in turn translates to an increase in the network throughput. Further, the upcoming MAC protocols for WLANs should work without message exchange between contenders, that is, work in a fully decentralised fashion in order to avoid injecting extra control traffic that may reduce the data throughput.

\textcolor{red}{Performing time slot reservation for each transmission is a well known technique for increasing the throughput and mantanining Quality of Service (QoS) in TDMA schemes, like LTE~\cite{canoLTEcoexistence}. Applying the same concept to CSMA networks by modifying DCF's random backoff proceedure provides similar benefits~\cite{HE}. The following are MAC protocols for WLANs, decentralised and capable of attaining greater throughput than CSMA/CA by constructing collision-free schedules using reservation techniques. A survey of collisio-free MAC protocols for WLANs is presented in~\cite{L_MAC}. In this paper we only overview those that are similar to CSMA/ECA.}

@@

Zero Collision MAC (ZC-MAC)~\cite{ZMAC} achieves a zero collision schedule for WLANs in a fully decentralised way. It does so by allowing contenders to reserve one empty slot from a predefined virtual schedule of $M$-slots in length. Backlogged stations pick a slot in the virtual cycle to attempt transmission. If two or more stations picked the same slot in the cycle, their transmissions will eventually collide. This forces the involved contenders to randomly and uniformly select other empty slot from those detected empty in the previous cycle plus the slot where they collided. When all $N$ stations reserve a different slot, a collision-free schedule is achieved.

ZC-MAC is able to outperform CSMA/CA under different scenarios. Nevertheless, given that the length of ZC MAC's virtual cycle has to be predefined without actual knowledge of the real number of contenders in the deployment, the protocol is unable to provide a collision-free schedule when $N>M$. Furthermore, if $M$ is overestimated ($M\gg N$), the fixed-width empty slots between each contender's successful transmission are no longer negligible and contribute to the degradation of the network performance. Additionally, Z-MAC nodes require common knowledge of where the virtual schedule starts/ends. This is not considered in CSMA/CA and constitutes an obstacle towards standardisation.

@@

Learning-MAC~\cite{L_MAC} is another MAC protocol able to build a collision-free schedule for many contenders. It does so defining a \emph{learning strength} parameter, $\beta\in(0,1)$. Each contender starts by picking a slot $s$ for transmission of the schedule $n$ of length $C$ at random with uniform probability. After a contender picks slot $s(n)$, its selection in the next schedule, $s(n+1)$, will be conditioned by the result of the current attempt. (\ref{success}) and (\ref{collisions-eq}) extracted from~\cite{L_MAC} show the probability of selecting the same slot $s(n)$ in cycle $n+1$.

@@

L-MAC is able to achieve higher throughput than CSMA/CA with a very fast convergence speed. Nevertheless, the choice of $\beta$ suppose a previous knowledge of the number of empty slots ($C-N$, where $N$ is the number of contenders), which is not easily available to CSMA/CA or may require a centralised entity~\cite{barcelo2011tcf}.

### 2015-10-29 results.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

\section{Results}\label{results}
	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/saturation-combined.eps}
		\caption{Saturation traffic results: a) Throughput under saturated conditions; b) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; c) Average percentage of collision slots: fraction of time slots containing collisions; d) Average time between successful transmissions (sx tx), e) Throughput when increasing the clock drift probability}
		\label{fig:satResults}
	\end{figure*}

@@

	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
		\caption{Saturation traffic results: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
		\label{fig:unsatResults}
	\end{figure*}

@@

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers layers must be discarded or blocked}
		\label{fig:blocked-packets}
	\end{figure}

@@

	\textcolor{red}{Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}). This is due to the construction of collision-free schedules which ensure that nodes transmit frequently (Hysteresis) and Fair Share which reduces the average delay given that more packets are transmitted and acknowledged with a single ACK (A-MPDU aggregation).}

@@

	\textcolor{red}{Figure~\ref{fig:coexResults}e shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA). It is possible to see that for a low total contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network. Again, this is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.}

	\textcolor{red}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only.}

@@

	\textcolor{red}{Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both \emph{SR aggr. halv. dynStick} curves (with and without Fair Share) have higher aggregated throughput than the rest. This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and mantains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.}

	\textcolor{red}{Figure~\ref{fig:SchedResetResults}c shows the average time between successful transmissions for the same network setup. Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost a $43\%$.}

	\textcolor{red}{Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregate througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$. Although a reduction in time between successfull transmissions is achieved with Schedule Reset, under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues. Therefore Schedule Reset's benefits are relevant only when in saturation.}

>>>>

\section{Results}\label{results}
	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/saturation-combined.eps}
		\caption{Simulation results under saturated traffic: a) Throughput under saturated conditions; b) Evolution of the fraction of collision slots in a scenario with 70 saturated stations; c) Average percentage of collision slots: fraction of time slots containing collisions; d) Average time between successful transmissions (sx tx), e) Throughput when increasing the clock drift probability}
		\label{fig:satResults}
	\end{figure*}

@@

	\begin{figure*}[tb]
		\centering
		\includegraphics[width=0.5\linewidth,angle=-90]{figures/tonFigs/nonsaturation-combined.eps}
		\caption{Simulation results under non-saturated traffic: a) Throughput; b) Average percentage of collision slots: the fraction of time slots containing collisions; c) Average time between successful transmissions; d) Average fraction of dropped packets; e) Average system delay; f) Average number of packets in the MAC queue of a node}
		\label{fig:unsatResults}
	\end{figure*}

@@

	\begin{figure}[tb]
		\centering
		\includegraphics[width=0.7\linewidth,angle=-90]{figures/unsaturated/blockingProb-unsat/blocking-unsaturated-TON.eps}
		\caption{Average fraction of blocked packets. When the MAC queue is full, packets comming from the uppers are discarded (or blocked) at the entrance of the buffer}
		\label{fig:blocked-packets}
	\end{figure}

@@

	\textcolor{red}{Further, with CSMA/ECA$_{\text{Hys+FS}}$ the percentage of blocked packets from the MAC queue is lower than CSMA/CA or CSMA/CA$_{\text{MaxAg}}$ (see Figure~\ref{fig:blocked-packets}). This is due to the construction of collision-free schedules which ensure that large A-MPDU transmissions do not suffer from collisions.}

@@

	\textcolor{red}{Figure~\ref{fig:coexResults}e shows the average throughput for CSMA/CA and CSMA/ECA$_{\text{Hys+FS}}$ stations in a saturated mixed setup (half the contenders using CSMA/CA). It is possible to see that for a low total number contenders ($N^{*}\leq 12$) CSMA/CA stations attain greater throughput than in a CSMA/CA-only network. Again, this is because in the mixed network setup the other $N^{*}/2$ contenders with CSMA/ECA$_{\text{Hys+FS}}$ use a deterministic backoff, leaving many empty slots between transmissions.}

	\textcolor{red}{Still referring to Figure~\ref{fig:coexResults}e, for $N^{*}>12$ periods of collision-free operation and the aggregation performed with Fair Share allows CSMA/ECA$_{\text{Hys+FS}}$ to have larger channel time than CSMA/CA nodes, which throughput degrades even more than in CSMA/CA-only. These results suggests that a switch to CSMA/ECA$_{\text{Hys+FS}}$ can be beneficial for networks with high number of contenders are expected.}

@@

	\textcolor{red}{Still assessing Figure~\ref{fig:SchedResetResults}a and focusing on the SR curves, we can see that both aggressive schedule halving with dynamic stickiness (\emph{SR aggr. halv. dynStick}) curves (with and without Fair Share) have higher aggregated throughput than the rest (for a description of aggressive schedule halving and dynamic stickiness, refer to Algorithm~\ref{alg:schedRest} and Section~\ref{aggr}). This Schedule Reset configuration reduces the time between successful transmissions (see Figure~\ref{fig:SchedResetResults}c) and mantains collision-free operation for longer periods of time thanks to the increase in the node's stickiness.}

	\textcolor{red}{Figure~\ref{fig:SchedResetResults}c shows the average time between successful transmissions for the same network setup. Even-though CSMA/ECA$_{\text{Hys+FS}}$ still has a higher metric than CSMA/CA due to the aggregation done by Fair Share, Schedule Reset is able of reducing this metric by almost $43\%$.}

	\textcolor{red}{Figure~\ref{fig:SchedResetResults}b and Figure~\ref{fig:SchedResetResults}d show the aggregated througput and time between successful transmissions, respectively for a non-saturated network with the same $p_e=0.1$. Since under non-saturated conditions CSMA/ECA$_{\text{Hys+FS}}$ nodes reset their schedule every time they empty their MAC queues, Schedule Reset's benefits are not relevant. Nevertheless, a reduction in the time between successful transmissions is observed for Schedule Reset curves in Figure~\ref{fig:SchedResetResults}d}.

### 2015-10-29 simulations.tex
Incorporated Francesco's and Boris's suggestions. Final Real pending

\section{Simulation Scenario}\label{simulations}
This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated. \textcolor{red}{We also provide details on how channel imperfections are modelled and what are its effects over the transmissions. Further, the simulation of the clock drift effect, Schedule Reset parameters, and the coexistence with CSMA/CA are also subjects to be addressed in this section.}

@@

	\begin{itemize}
		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) standard.
		\item All nodes are in communication range.
		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
		\item Collisions take as much channel time as successful transmissions.
	\end{itemize}

@@

	\begin{table}
		\centering
		\caption{PHY and MAC parameters for the simulations \textcolor{red}{REVIEW}}
		\label{tab:mac-params}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 65~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Data payload (Bytes) & 1024\\
			MAC queue size (Packets) & 1000\\
			\hline
		\end{tabular}
	\end{table}

@@

	\textcolor{red}{\subsection{Channel errors}\label{channelErrorsDef}
	Channel errors are modelled by assigning a probability of a packet being corrupted by the channel, $p_e>0$. That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged. If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each MPDU individually and independently. Therefore, an A-MPDU transmission will be considered a failure if all frames in the aggregation are affected by the channel error probability.}

>>>>

\section{Simulation Scenario}\label{simulations}
This section provides the simulation parameters for testing CSMA/ECA$_{\text{Hys+FS}}$ under two different traffic conditions, namely saturated and non-saturated. \textcolor{red}{We also provide details on how channel errors are modelled and what are its effects over the transmissions. Further, the simulation of the clock drift effect, Schedule Reset parameters, and the coexistence with CSMA/CA are also subjects to be addressed in this section.}

@@

	\begin{itemize}
		\item Unspecified parameters follow the IEEE 802.11n ($2.4$~GHz) amendment.
		\item All nodes are in communication range.
		\item No Request-to-Send (RTS) or Clear-to-Send (CTS) messages are used.
		\item Collisions take as much channel time as successful transmissions.
	\end{itemize}

@@

	\begin{table}
		\centering
		\caption{PHY and MAC parameters for the simulations}
		\label{tab:mac-params}
		\begin{tabular}{|c|c|}
			\hline
			\multicolumn{2}{|c|}{{\bfseries PHY}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			PHY rate & 65~Mbps\\
			Empty slot & $9~\mu s$\\
			DIFS & $28~\mu s$\\
			SIFS & $10~\mu s$\\
			\hline
			\multicolumn{2}{|c|}{{\bfseries MAC}}\\
			\hline
			{\bfseries Parameter} & {\bfseries Value}\\
			\hline
			Maximum backoff stage ($m$) & 5\\
			Minium Contention Window ($CW_{\min}$) & 16\\
			Maximum retransmission attempts & 6\\
			Data payload (Bytes) & 1024\\
			MAC queue size (Packets) & 1000\\
			\hline
		\end{tabular}
	\end{table}

@@

	\textcolor{red}{\subsection{Channel errors}\label{channelErrorsDef}
	Channel errors are modelled by assigning a probability of a packet being corrupted by the channel, $p_e>0$. That is, in a single packet transmission there is probability $p_e$ that the transmission will not be acknowledged. If the transmission is an A-MPDU (like in the case of CSMA/ECA$_{\text{Hys+FS}}$), $p_e$ will affect each MPDU individually and independently. Therefore, an A-MPDU transmission will be considered a complete failure only if all frames in the aggregation are affected by the channel error probability.}
